---
title: "05_createstudyarea"
author: "Michaela Gustafson"
date: "2022-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Create Gyrfalcon Study Area

### Library
```{r lib}
library(terra)
library(sf)
library(sp)
library(rgdal)
library(ggplot2)
library(tidyverse)
library(raster)
library(tidyr)
library(dplyr)
library(stringr)
library(rgeos)
library(here)
```


### Import point count shapefile and occupancy data (what I've got for now) to create shapefile for GYRF territories


```{r import}
pts <- readOGR(here("shapefiles"), "pts_3years")

bente <- read.csv(here("data/BenteGyrfalonTerritorySurveys.csv"))

ak_dem <- raster::raster(here("data/dsm_mosaic_AK.tif"))
```


Subset Bente data to be Site (which is the territory level) and coordinates:

```{r subbente}
colnames(bente)
bente.sites <- bente[,c(2, 4:5, 26)]# site:lat:lon:year
```

# For sites with repeating years, only need to keep one occurrence:

```{r duplicates}
unique(bente.sites$site)
# should end up with 271 rows because 271 unique GYRF territories

bente.sub <- bente.sites %>% distinct(site, .keep_all = TRUE)

```


# Now turn coordinates into shapefile and put in same CRS as elevation raster

```{r shpfle}
xy4 <- bente.sub[c(3,2)]
colnames(xy4)[1] <- "x"; colnames(xy4)[2] <- "y"
coordinates(xy4) <- ~ x + y
spdf2 <- SpatialPointsDataFrame(coords = xy4, data = bente.sub)
proj4string(spdf2) <- CRS("+init=epsg:4326")

bente.proj <- spTransform(spdf2, crs(ak_dem))
#plot(ak_dem); points(bente.proj)
writeOGR(bente.proj, dsn=here("shapefiles"), layer="bente_proj_2" , driver="ESRI Shapefile")

```

### Buffer points

Okay, I need to buffer territories with a 15km buffer +500m (15500 m) 

And buffer survey points by 500m *** HAD TO CHANGE TO 4000m buffer so that no gaps in study area polygon

Then dissolve polygon buffers together

```{r buffer}

# GYRF Territories:

# check what units of measurement we are in so we know what to put for width:
st_crs(bente.proj, parameters = TRUE)$units_gdal

# create buffer
bente.buff <- gBuffer(bente.proj, byid = TRUE, width = 15500)
plot(bente.buff)


# Prey Survey Points
# check what units of measurement we are in so we know what to put for width:
st_crs(pts, parameters = TRUE)$units_gdal

# create buffer
pts.buff <- gBuffer(pts, byid = TRUE, width = 4000)
plot(pts.buff)

# Merge buffers
study.area <- raster::bind(bente.buff, pts.buff)
plot(study.area)


# Dissolve merged shapefiles

study.diss <- aggregate(study.area)
plot(study.diss)
proj4string(study.diss)

shapefile(x = study.diss, file = "shapefiles/gyrf_studyarea.shp")

```

