---
title: "05_landvariables"
author: "Michaela Gustafson"
date: "2022-08-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In this script I will be extracting land cover variables, mean elevation, and range of elevation ('steepness') for each point count survey point.

I will be adding land cover raster data from: https://daac.ornl.gov/ABOVE/guides/Annual_Landcover_ABoVE.html

I will be importing GPS points for 2019 and 2021 and 2022 prey surveys


## LIBRARY

```{r library}
library(terra)
library(sf)
library(sp)
library(rgdal)
library(ggplot2)
library(tidyverse)
library(raster)
library(tidyr)
library(dplyr)
library(stringr)
library(rgeos)
library(here)
library(reshape2)
```



## IMPORT DEM
Import elevation raster 
```{r dem}
ak_dem <- raster::raster("E:/gyrf_analysis/gyrf3/data/dsm_mosaic_AK.tif")
```
# Also - we want to put all other rasters into the same crs as the digital elevation model

## UPLOAD LAND COVER RASTERS
Uploading the raster that Jenny ran for me that has the land cover tiles mosaiced and in the same CRS as the DEM

```{r landcover}

### JENNY'S CODE:
#all.mos <- raster("E:/Dropbox/Back&Forth/GRYF/michaela/ABoVE_2014lyrs/all.mos.tif")
#plot(all.mos)

all.mos.dem <- projectRaster(all.mos, method="ngb", crs = "+proj=aea +lat_0=50 +lon_0=-154 +lat_1=55 +lat_2=65 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs")
#writeRaster(all.mos.dem, "E:/Dropbox/Back&Forth/GRYF/michaela/ABoVE_2014lyrs/all.mos.dem.tif", overwrite=TRUE)

test.mos <- raster("E:/gyrf_analysis/gyrf3/data/ABoVE_2014lyrs/all.mos.tif")
res(test.mos)

test.mos.in.dem <- projectRaster(test.mos, method="ngb", crs = "+proj=aea +lat_0=50 +lon_0=-154 +lat_1=55 +lat_2=65 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs")

res(test.mos.in.dem)

all.mos <- raster("E:/gyrf_analysis/gyrf3/jenny/all.mos.dem.tif")
res(all.mos)

save.image("landvariables_wkspce.Rdata")
```

## IMPORT SURVEY LOCATIONS
Already compiled these and put in the correct CRS in previous code

```{r pionts}
# these points were previously put in same CRS as dem
pts <- as.data.frame(readOGR("E:/gyrf_analysis/gyrf3/shapefiles","pts_3years_2"))

```


## Study Area

```{r studyarea}
study.area <- readOGR("E:/gyrf_analysis/gyrf3/shapefiles", "gyrf_studyarea")
crs(study.area)
crs(ak_dem)
study.area.dem <- spTransform(study.area, crs(ak_dem))

res(study.area.dem)
```



Step 1: 
a. Crop and mask elevation raster with study area.
b. Aggregate elevation raster from 5m res to 400m res
c. Remove pixels that result in a mean elevation >500m

```{r akdem}
# Step A: Crop and mask elevation raster with study area
ak_dem.crop <- raster::crop(ak_dem, study.area.dem)
ak_dem.crop.mask <- raster::mask(ak_dem.crop, study.area.dem)

res(ak_dem.crop.mask)

# Step B: Aggregate elevation raster from 5m res to 400m res
dem.400 <- aggregate(ak_dem.crop.mask, fact = 80, fun = mean, na.rm = TRUE) # have to include na.rm = true or else elevations >500m come back into the picture!

# Step C: remove pixels that result in mean elevation >500m
dem.400[dem.400>=500] <- NA


# save for easier code running later
writeRaster(dem.400, here("output/dem_400.tif"), overwrite = TRUE)

# Here I saved and closed the script and project and restarted then just loaded the dem_400.tif to clear out some memory

dem.400 <- raster(here("output/dem_400.tif"))
```


## Crop land cover to study area then resample to final elevation raster

```{r}

mos.crop <- raster::crop(all.mos, study.area.dem)
mos.mask <- raster::mask(mos.crop, study.area.dem)


## Run code to aggregate to 400 and get percentages of land cover types and creating prediction surface


r <- raster(extent(dem.400), res = 400, crs = "+proj=aea +lat_0=50 +lon_0=-154 +lat_1=55 +lat_2=65 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs ")

############# prediction surface land cover ###############
neighborhood_radius <- 400 / 2

# get cell centers and create neighborhoods
r_centers <- rasterToPoints(r, spatial = TRUE) %>% 
  st_as_sf() %>% 
  transmute(id = row_number())
r_cells <- st_buffer(r_centers, dist = neighborhood_radius)

# extract land cover values within neighborhoods
lc_extract_pred <- mos.mask %>% 
  exactextractr::exact_extract(r_cells, progress = TRUE) %>% 
  map(~ count(., mos.mask = value)) %>% 
  tibble(id = r_cells$id, data = .) %>% 
  unnest(data)

# calculate the percent for each land cover class
pland_pred <- lc_extract_pred %>% 
  count(id, mos.mask) %>% 
  group_by(id) %>% 
  mutate(pland = n / sum(n)) %>% 
  ungroup() %>% 
  dplyr::select(-n) %>% 
  # remove NAs after tallying so pland is relative to total number of cells
  filter(!is.na(mos.mask))

lc_names <- tibble(mos.mask = 1:15,
                   lc_name = c("egforest", "decforest", "mixforest", "woodland", "lowshrub", "tallshrub", "openshrub", "tundra", "tussock", "sparseveg", "fen", "bog", "littoral", "barren", "water"))

# convert names to be more descriptive
pland_pred <- pland_pred %>% 
  inner_join(lc_names, by = "mos.mask") %>% 
  arrange(mos.mask) %>% 
  dplyr::select(-mos.mask)


# tranform to wide format, filling in implicit missing values with 0s
pland_pred <- pland_pred %>% 
  pivot_wider(names_from = lc_name, 
              values_from = pland, 
              values_fill = list(pland = 0)) %>% 
  dplyr::select(id, everything())

# join in coordinates
pland_coords <- st_transform(r_centers, crs = "+proj=aea +lat_0=50 +lon_0=-154 +lat_1=55 +lat_2=65 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs") %>% 
  st_coordinates() %>% 
  as.data.frame() %>% 
  cbind(id = r_centers$id, .) %>% 
  rename(longitude = X, latitude = Y) %>% 
  inner_join(pland_pred, by = "id")

lc_list <- list()

for(i in 1:length(lc_names)){
lc_list <- pland_coords %>% 
  # convert to spatial features
  st_as_sf(coords = c("longitude", "latitude"), crs = "+proj=aea +lat_0=50 +lon_0=-154 +lat_1=55 +lat_2=65 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs") %>% 
  st_transform(crs = projection(r)) %>% 
  # rasterize points
   rasterize(r, field = lc_names[[i]]) %>% 
  # trim off empty edges of raster
  trim()
}

lc_stack2 <- stack(lc_list)

#mask with dem.400
lc_stack2 <- resample(lc_stack2, dem.400, method = 'ngb') #NOT SURE WHY I DIDN'T DO THIS HERE? Earlier it was hashed out..
lc_stack2 <- mask(lc_stack2, dem.400)

writeRaster(lc_stack2, filename=names(lc_stack2), bylayer=TRUE, format="GTiff", overwrite = TRUE)

############################################################################
######## extract to the survey points from original res rasters ############
############################################################################
################## START HERE AUG 9th ###########################

############################################################################
######## extract to the survey points from original res rasters ############
############################################################################
colnames(pts)[3] <- "longitude"; colnames(pts)[4] <- "latitude"
neighborhood_radius <- 400 / 2

pts_buff <- pts[c(1,3,4)] %>% 
  # convert to spatial features
  st_as_sf(coords = c("longitude", "latitude"), crs = "+proj=aea +lat_0=50 +lon_0=-154 +lat_1=55 +lat_2=65 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs ") %>% 
  # buffer to create neighborhood around each point
  st_buffer(dist = neighborhood_radius) 


# iterate over all years extracting land cover for all pts buffer in each



plot(mos.mask)

lc_extract <- mos.mask %>% 
  exactextractr::exact_extract(pts_buff, progress = TRUE) %>% 
  map(~ dplyr::count(., lc = value)) %>%
  tibble(id = pts_buff$id, data = .) %>%
  unnest(data)

pland <- lc_extract %>% 
  # calculate proportion
  dplyr::group_by(id) %>% 
  dplyr::mutate(pland = n / sum(n)) %>% 
  ungroup() %>% 
  dplyr::select(-n) %>% 
  # remove NAs after tallying so pland is relative to total number of cells
  dplyr::filter(!is.na(lc))

lc_names <- tibble(lc = 1:15,
                   lc_name = c("egforest", "decforest", "mixforest", "woodland", "lowshrub", "tallshrub", "openshrub", "tundra", "tussock", "sparseveg", "fen", "bog", "littoral", "barren", "water"))

pland <- pland %>% 
  inner_join(lc_names, by = "lc") %>% 
  arrange(lc) %>% 
  dplyr::select(-lc)

# transform to wide format, filling in implicit missing values with 0s%>% 

colnames(pland)[2] <- "value"
pland$lc_name <- as.factor(pland$lc_name) #need to be a factor not character

pland2 <- reshape2::dcast(pland, id~lc_name, sum) # needs a function, shouldn't matter if its mean or sum b/c  there is only one per id/ lc type
pland2[is.na(pland2)] <- 0 # fill in NAs with 0

#
########################################################################
#pland <- pland %>%
 # pivot_wider(names_from = lc_name, 
              #values_from = pland, 
              #values_fill = list(pland = 0))
              



#pland3 <- pland[,-c(2)]
#pland2 <- aggregate(x = pland3, by = list(id), FUN = "sum")
######## extract dem: using extract() here - it takes longer for sure! #####

#ak_dem <- raster("E:/Dropbox/Back&Forth/GRYF/michaela/ak_dem.tif") # already loaded earlier
###################################################################################################
mean.elev <- as.data.frame(extract(ak_dem, pts_buff, fun=mean, na.rm=TRUE))
colnames(mean.elev)[1] <- "elev"

steep.elev <- as.data.frame(extract(ak_dem, pts_buff, fun=range, na.rm=TRUE))
steep.elev$steep <- (steep.elev$V2) - (steep.elev$V1)

### put the covs together and export #####
data <- cbind(pland2, mean.elev)
data2 <- cbind(data, steep.elev[,3])
colnames(data2)[17] <- "steep"


write.csv(data2, here("output/data_pts_17Aug.csv"), row.names = FALSE)
```



### END SCRIPT





























