---
title: "10_ropt"
author: "Michaela Gustafson"
date: "1/6/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Paper: Modelling habitat suitability for alpine rock ptarmigan (Lagopus muta helvetica)
combining object-based classification of IKONOS imagery and Habitat Suitability
Index modelling
Margit Zohmanna, Josef Pennerstorfer, Ursula Nopp-Mayr

Rock Ptarmigan prefer more alpine habitats - rock/scree, rock with sparse vegetation, dwarf shrubs, and home ranges tend to be large because suitable habitat is patchy (this is from another paper: Spacing behaviour and habitat use of rock ptarmigan
(Lagopus mutus) at low density in the Italian Alps
Massimo Favaron, Giovanni C. Scherini,
Damiano Preatoni, Guido Tosi, Luc A. Wauters)

So maybe for ropt a little different: low shrub, sparseveg, barren

## TIME REMOVAL ANALYSIS: ROCK PTARMIGAN

In this code I will be running my time-removal abundance model for Rock Ptarmigan.

This is a multinomial N-mixture model.

## LIBRARY
These are the packages used in the following code:
```{r library}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(unmarked)
library(AICcmodavg)
library(raster)
library(here)
library(AHMbook)
library(brms)
```


## LOAD OBSERVATION and ENVIRONMENTAL DATA

```{r loaddata}
all.obs <- read.csv(here("output/obs_3years.csv"))
all.env2 <- read.csv(here("output/final_env_join_22Aug22.csv"))

# remove extra 'X' column from import
all.obs <- all.obs[,-c(1)]
all.env2 <- all.env2[,-c(1)]

for (i in 1:nrow(all.env2)) {
   all.env2$shrub[i] <- all.env2$openshrub[i] + all.env2$tallshrub[i]
}

for (i in 1:nrow(all.env2)) {
   all.env2$allshrub[i] <- all.env2$openshrub[i] + all.env2$tallshrub[i] + all.env2$lowshrub[i]
}

```
## PREP OBSERVATION DATA FOR SELECTED SPECIES

```{r roptprep}
# copy data table for WIPT

ropt <- data.frame(all.obs)

# change all species codes of WIPT to a 1 and all other species to 0
ropt$species[ropt$species == "ROPT"] <- 1
ropt$species[ropt$species != "1"] <- 0

# change count to 0 for non ROPT
ropt$count[ropt$species == 0] <- 0
head(ropt)
# now 'ropt' df represents counts for Rock Ptarmigan


# check for na values
sum(is.na(ropt$time_int))


# change time intervals to reflect intervals of 2 minutes
ropt2 <- mutate(ropt, 
                time_int = ifelse(ropt$time_int %in% 0:1, "1", 
                                  ifelse(ropt$time_int %in% 2:3, "2", 
                                         ifelse(ropt$time_int %in% 4:5, "3",
                                                ifelse(ropt$time_int %in% 6:7, "4",
                                                       ifelse(ropt$time_int %in% 8:9, "5", "NA"))))))


# aggregate rows and sum counts for matching keys and time intervals
# must change formats first:

str(ropt2)

ropt2$id <- as.factor(ropt2$id)
ropt2$time_int <- as.factor(ropt2$time_int)
ropt2$count <- as.integer(ropt2$count)


sum(ropt2$count) # sample size 73

ropt.agg <- aggregate(x = ropt2$count, 
                      by = list(ropt2$id, ropt2$time_int), 
                      FUN = sum)




# rename columns in aggregated df

head(ropt.agg)

names(ropt.agg)[names(ropt.agg) == "Group.1"] <- "id"
names(ropt.agg)[names(ropt.agg) == "Group.2"] <- "time_int"
names(ropt.agg)[names(ropt.agg) == "x"] <- "count"

head(ropt.agg)

# okay so ropt.agg is our count dataframe, we don't need any of the other columns,
# those were only used to help filter out for distance and flyover observatiosn

# check that ids are matching between env and obs
# need to change id to a factor to use anti-join
all.env2$id <- as.factor(all.env2$id)
miss1 <- anti_join(all.env2, ropt.agg) # clear
miss2 <- anti_join(ropt.agg, all.env2) # clear

# nothing there so that's good

# spread dataframes:

unique(ropt.agg$id) # should end up with 988 rows


ropt.wide <- ropt.agg %>%
  dplyr::select(id, time_int, count) %>%
  spread(key = time_int, value = count, fill = 0)

write.csv(ropt.wide, here("output/ropt_wide_22aug22.csv"))
ropt.wide <- read.csv(here("output/ropt_wide_22aug22.csv"))
ropt.wide <- ropt.wide[,-c(1)]

```


# Test for variable outliers

```{r corrtest}

test.df <- left_join(ropt.wide, all.env2)
test.df$total <- rowSums(test.df[,c("1", "2", "3", "4", "5")])

a1 <- ggplot( data = test.df) + geom_histogram( aes_string( fill = (as.factor(test.df$total)), x = test.df$sparseveg))
a1 # for ROPT the upper ones might be outliers...above 0.7-75ish?

a2 <- ggplot( data = test.df) + geom_histogram( aes_string( fill = (as.factor(test.df$total)), x = test.df$tundra))
a2 #

a4 <- ggplot( data = test.df) + geom_histogram( aes_string( fill = (as.factor(test.df$total)), x = test.df$barren))
a4 # possible outliers here as well around 0.35

a5 <- ggplot( data = test.df) + geom_histogram( aes_string( fill = (as.factor(test.df$total)), x = test.df$lowshrub))
a5 # outliers here for low shrub




#okay might have to use all shrub since low shrub has some outliers? or could just use shrub: tall and open together...




# sparseveg is out
# tundra is okay


# So ROPT model might be:
# tundra
# shrub
# elev


```









## SCALE AND CHECK CORRELATIONS

```{r scaledf}

all.env.scaled2 <- read.csv(here("output/all.env.scaled_22Aug22.csv"))
all.env.scaled2 <- all.env.scaled2[,-c(1)]
```

## TIME-REMOVAL MODEL
WIPT sample size: 62 individuals counted
```{r trmodel}

# check structures and join data frames
colnames(all.env.scaled2)
str(all.env.scaled2)
str(ropt.wide)

ropt.env.join <- left_join(ropt.wide, all.env.scaled2)
head(ropt.env.join); dim(ropt.env.join)
str(ropt.env.join)
ropt.env.join$year <- as.factor(ropt.env.join$year)

# create observation and site covariate data frames
colnames(ropt.env.join)

timeints <- ropt.env.join[,c(2:6)]
siCovs <- ropt.env.join[,c(7:32)]
area <- pi*400^2


roptFrame2 <- unmarkedFrameMPois(
  # import time removal columns(counts):
  y = timeints, 
  siteCovs = data.frame(siCovs, area),
  # define pifun type: 
  type = "removal" )

# fit models: multinomPois order of formulas: detection, abundance
# I will only be running a null model and a full model. Not gonna mess around with that multiple model bullshit
# can't use all land cover variables because htey add up to 1 within each site

### ADD WIND BACK IN??? AND DEM FOR TEST

fm0 <- multinomPois(~ 1 ~ 1, data = roptFrame2) #null model
fm.1 <- multinomPois( ~ 1 + julian + min_after_sun + wind + observer ~ 1 + tundra + shrub + elev + offset(log(area)), data = roptFrame2)
fm.10 <- multinomPois( ~ 1 + julian + min_after_sun + wind + observer ~ 1 + tundra + lowshrub + sparseveg  + elev + offset(log(area)), data = roptFrame2)
fm.11 <- multinomPois( ~ 1 + julian + min_after_sun + wind + observer ~ 1 + lowshrub + sparseveg + barren + elev + offset(log(area)), data = roptFrame2)


ms2 <- fitList(
  "lam(.)p(.)" = fm0,
  "lam(tundra+shrub+elev+offset(log(area)))p(julian+min_after_sun+wind+observer)" = fm.1,
  "lam(tundra+lowshrub+sparseveg+elev+offset(log(area)))p(julian+min_after_sun+wind+observer)" = fm.10,
  "lam(lowshrub+sparseveg+barren+elev+offset(log(area)))p(julian+min_after_sun+wind+observer)" = fm.11)

(ms2sel <- modSel(ms2))
ms2sel


summary(fm0)
summary(fm.1)

summary(fm.10)
summary(fm.11)
# does the model fit okay: compare model with intercept only [anova] look at AIC
# when drop out cats, compare AIC (if makes big difference might have to put back in?)

# residual plots

### nonsignificant cat variables
## hear? try removing and check other values maybe observer? b/c of std error

```

## CHECK MODEL FIT
Will be looking at c-hat for overdispersion, as well as chisq value and its p-value
```{r gfone, eval=FALSE}
set.seed(10011)
#NULL
fitstats.null <- function(fm0) {
  observed <- getY(fm0@data)
  expected <- fitted(fm0)
  resids <- residuals(fm0)
  sse <- sum(resids^2)
  chisq <- sum((observed - expected)^2 / expected)
  freeTuke <- sum((sqrt(observed) - sqrt(expected))^2)
  out <- c(SSE=sse, Chisq=chisq, freemanTukey=freeTuke)
  return(out)
}


(gof.fm.null <- parboot(fm0, fitstats.null, nsim = 1000, report = 1))

# FULL

fitstats.10 <- function(fm.10) {
  observed <- getY(fm.10@data)
  expected <- fitted(fm.10)
  resids <- residuals(fm.10)
  n.obs <- apply(observed,1,sum,na.rm=TRUE)
  n.pred <- apply(expected,1,sum,na.rm=TRUE)
  sse <- sum(resids^2, na.rm=TRUE)
  chisq <- sum((observed - expected)^2 / expected, na.rm=TRUE)
  freeTuke <- sum((sqrt(observed) - sqrt(expected))^2, na.rm=TRUE)
  freeTuke.n <- sum((sqrt(n.obs)-sqrt(n.pred))^2, na.rm=TRUE)
  sse.n <- sum( (n.obs-n.pred)^2, na.rm=TRUE)
  chisq.n <- sum((n.obs-n.pred)^2/expected, na.rm=TRUE)
  
  out <- c(SSE=sse, Chisq=chisq, freemanTukey=freeTuke, SSE.n=sse.n, Chisq.n=chisq.n, freemanTukey.n=freeTuke.n)
  return(out)
}

(gof.fm.10 <- parboot(fm.10, fitstats2, nsim = 1000, report = 1))

print(gof.fm.null)
print(gof.fm.10)

# compute c-hat: 0.9398  but model not so good...
chat.10 <- gof.fm.10@t0[2]/mean(gof.fm.10@t.star[,2])
chat.10

saveRDS(fm.10, here("models/roptfm_15oct22.rds"))
fm.10 <- readRDS(here("models/roptfm_15oct22.rds"))


rp.coef <- as.data.frame(coef(fm.10))
rp.state.confint <- as.data.frame(confint(fm.10, level = 0.95, type = "state"))
rp.det.confint <- as.data.frame(confint(fm.10, level = 0.95, type = "det"))
write.csv(rp.coef, here("output/rpcoef.csv"))
write.csv(rp.state.confint, here("output/rpci_state.csv"))
write.csv(rp.det.confint, here("output/rpci_det.csv"))


save.image(here("10_ropt_workspace.RData"))
```


## PREDICT!

```{r predict}

# First need to scale study-wide rasters to match scaling of model:


elev <- raster("E:/gyrf_analysis/gyrf3/output/dem_400.tif")
lowshrub <- raster("E:/gyrf_analysis/gyrf3/prediction surfaces 17Aug/lowshrub.tif")
sparseveg <- raster("E:/gyrf_analysis/gyrf3/prediction surfaces 17Aug/sparseveg.tif")
tundra <- raster("E:/gyrf_analysis/gyrf3/prediction surfaces 17Aug/tundra.tif")


area.raster <- elev 
values(area.raster) <- 400*400 # area of a grid pixel

# find center/mean and std dev of scaled covariates to scale prediction surfaces above

elev.s <- (elev-185)/106

lowshrub.s <- (lowshrub-0.0562)/0.0794

sparseveg.s <- (sparseveg-0.0818)/0.168

tundra.s <- (tundra-0.57)/0.306

# add detection covariates (standardized mean values)

jul.raster <- elev 
values(jul.raster) <- mean(all.env.scaled2$julian)

wind.raster <- elev 
values(wind.raster) <- mean(all.env.scaled2$wind)

minsun.raster <- elev 
values(minsun.raster) <- mean(all.env.scaled2$min_after_sun)

obs.raster <- elev 
values(obs.raster) <- as.factor("DS")


pred.surface <- stack(elev.s, lowshrub.s, sparseveg.s, tundra.s, area.raster, jul.raster, wind.raster, minsun.raster, obs.raster)

names(pred.surface) <- c("elev", "lowshrub", "sparseveg", "tundra", "area", "julian", "wind", "min_after_sun", "observer")


ropt.prediction <- predict(fm.10, type='state', newdata=pred.surface)

ropt.freq <- freq(ropt.prediction, digits = 2)

ropt.freq1 <- as.data.frame(ropt.freq[1])

write.csv(ropt.freq1, here("output/roptcellcounts.csv"))


writeRaster(ropt.prediction, filename = names(ropt.prediction), bylayer = TRUE, format = "GTiff", overwrite = TRUE)


```



# PREDICTION MAP

```{r roptpredmap}

plot(ropt.prediction, axes=FALSE, col=terrain.colors(100))

sum_raster_cells <-cellStats(ropt.prediction, 'sum')
sum_raster_cells


```


# Out of sample prediction

This is a pretty big dataset (for ecology, at least), so if we might choose to do strict out-of-sample prediction, in which we "hold out" 1/4 or 1/3 of our data as a "test" dataset. We will fit the model on the "training" dataset and then assess out-of-sample prediction using the test:

```{r}

colnames(ropt.wide)
colnames(all.env.scaled2)
full.data <- merge(ropt.wide, all.env.scaled2)

full.data <- full.data[,c(1:6, 17, 12, 15, 21)]
full.data$area <- pi*400^2

fitted_value <- data.frame(nrows = 988, ncol = 4)

for(i in 1:nrow(full.data)) { 
    
   data_one_row_out<-full.data[-i,] #leaving one row of data out
  data_to_predict<-full.data[i,] #only the row you are predicting

  timeints <- data_one_row_out[,c(2:6)]
  siCovs <- data_one_row_out[,c(7:11)]

  one_out_frame <- unmarkedFrameMPois(
  # import time removal columns(counts):
  y = timeints, 
  siteCovs = siCovs,
  # define pifun type: 
  type = "removal" )
  
  model_fit<-multinomPois( ~ 1 ~ 1 + tundra + lowshrub + sparseveg  + elev + offset(log(area)), data=one_out_frame)# fitting the data to the one row out set
  
  fitted_value[i]<-predict(model_fit, type = 'state', newdata=data_to_predict[,c(7:11)])
  
}


```


```{r ppp}
# Rock Ptarmigan PPP

# Estimate partial prediction plots for predictors with 95% CIs not overlapping zero:
# Start by creating our datasets to predict over

# how many values do we use:
n <- 100

# Use the observed values to define our range:

lowshrub.pp <- seq( min( all.env2[,"lowshrub"]),max( all.env2[,"lowshrub"]),
                   length.out = n )

sparseveg.pp <- seq( min( all.env2[,"sparseveg"]),max( all.env2[,"sparseveg"]),
                   length.out = n )

#standardize predictors:
lowshrub.std <- scale( lowshrub.pp )
sparseveg.std <- scale( sparseveg.pp )
area.x = 400*400

```

```{r}
### Low shrub

#combine standardized predictors into a new dataframe to predict partial relationship
# for abundance submodel:
abundData <- data.frame( tundra = 0, lowshrub = lowshrub.std, sparseveg = 0, elev = 0, area = area.x )

#predict partial relationship:
pred.lows <- predict( fm.10, type = "state", newdata = abundData, 
                          appendData = TRUE )
#view
head( pred.lows ); dim( pred.lows )


### plot

r.lowsp <- cbind( pred.lows[,c("Predicted", "lower", "upper") ], lowshrub.pp ) %>%
  # define x and y values
  ggplot(., aes( x = lowshrub.pp, y = Predicted ) ) + 
  #choose preset look
  ylim(0, 0.2) +
  theme_bw( base_size = 50 ) +
  # add labels
  labs( x = "Low shrub (%)", y = "Rock Ptarmigan \n relative abundance" ) +
  # add band of confidence intervals
  theme(text=element_text(size=60)) +
  geom_smooth( aes(ymin = lower, ymax = upper ), 
               stat = "identity",
               size = 1.5, alpha = 0.5, color = "grey" ) +
  # add mean line on top
  geom_line( size = 2 ) 
#view
r.lowsp
ggsave("RockXLow2.png", width = 7, height = 7,dpi = 300)


```


```{r spveg}

#combine standardized predictors into a new dataframe to predict partial relationship
# for abundance submodel:
abundData <- data.frame( tundra = 0, lowshrub = 0, sparseveg = sparseveg.std, elev = 0, area = area.x )

#predict partial relationship:
pred.sv <- predict( fm.10, type = "state", newdata = abundData, 
                          appendData = TRUE )
#view
head( pred.sv ); dim( pred.sv )


### plot

r.svp <- cbind( pred.sv[,c("Predicted", "lower", "upper") ], sparseveg.pp ) %>%
  # define x and y values
  ggplot(., aes( x = sparseveg.pp, y = Predicted ) ) + 
  #choose preset look
  #ylim(0, 0.2) +
  theme_bw( base_size = 50 ) +
  # add labels
  labs( x = "Sparse veg (%)", y = "Rock Ptarmigan \n relative abundance" ) +
  # add band of confidence intervals
  #theme(axis.title.y=element_blank()) +
  theme(text=element_text(size=60)) +
  geom_smooth( aes(ymin = lower, ymax = upper ), 
               stat = "identity",
               size = 1.5, alpha = 0.5, color = "grey" ) +
  # add mean line on top
  geom_line( size = 2 ) 

r.svp

ggsave("RockXSparse2.png", width = 7, height = 7,dpi = 300)



```

```{r}
figure2 <- ggarrange(lowsp, svp,
                    ncol = 2, nrow = 1)


figure2
ggsave("roptfigs.png", width = 30, height = 10, dpi = 300, limitsize = FALSE)

save.image(here("10_ropt_workspace.RData"))

```




```{r}
library(ggpubr)

figure3 <- ggarrange(r.lowsp, r.svp, w.tundp, w.allsp, w.tussp, w.elevp, s.lowsp, s.svp, s.elevp,
                    ncol = 3, nrow = 3)

figure3

ggsave("allspfigs.png", width = 60, height = 40, dpi = 300, limitsize = FALSE)

```




